Relational programming in Smalltalk

### 2019-03-17 booklet/article.tex
a first pass

\documentclass[a4paper,12pt]{article}

@@

@@

\subsection{\texttt{Goal}s and \texttt{State}s}

A \textit{goal} is any object that responds to the selector \Verb|#onState:| so
that it receives a substitution and produces an ordered sequence of zero or
more substitutions. In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\Verb|Dictionary| object fits perfectly; we wrap it in a \Verb|State| object
that also keeps record of an incremental numbering \Verb|birthdate| whose
meaning will be clarified later when we'll describe the \Verb|Fresh|
goal\footnote{put here the reference to its subsection}. An empty substitution
is created by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary new asState
\end{minted}
where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary>>#asState
    ^ State new
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#at: aVar put: aValue
    | s |
    s := substitution copy.
    s "check that if a key already exists at aVar, then the values must be the same"
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
mantaining a coherent protocol with respect to the one of \Verb|Dictionary|
objects for extension; moreover, \Verb|UnificationError| objects will be
explained in a later section when will talk about the unification process.

\subsection{\texttt{Chain}s and \texttt{Cons}es}

Because the sequence of substitutions may be infinite, we
represent it not as a \Verb|LinkedList| object but as a \Verb|Chain| object;
as usual, we find \Verb|BlockClosure| objects good candidates to represent (a possibly
infinite) computation. Some examples are in order,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#collatz: o
    ^ o
        cons: [o even
                ifTrue: [self collatz: o / 2]
                ifFalse: [self collatz: 3 * o + 1]]

ChainTest>>#fibs
    ^ self fib: 0 fib: 1

ChainTest>>#fib: m fib: n
    ^ m cons: [self fib: n fib: m + n]

ChainTest>>#gcd: m gcd: n
    ^ m cons: [n isZero
                ifTrue: [Chain with: m]
                ifFalse: [self gcd: n gcd: m \\ n]]

ChainTest>>#nats
    ^ self ints: 0

ChainTest>>#ints: i
    ^ i cons: [self ints: i + 1]

ChainTest>>#zeroOneTwo
    ^ 0 cons: (1 cons: (2 cons: Chain bottom))
\end{minted}
where any \Verb|Object| object respond to the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#cons: anObj
    ^ anObj consedObject: self
\end{minted}
which dispatches over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#consedObject: car
    ^ Cons car: car asCons cdr: self asCons

BlockClosure>>#consedObject: car
    ^ Chain item: car linker: self

Chain>>#consedObject: car
    ^ car cons: [self]
\end{minted}
where (i)~\Verb|Cons| objects mimic \textit{Lisp's cons cells}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons class>>#car: anObj cdr: anotherObj
    ^ self new
        car: anObj;
        cdr: anotherObj;
        yourself

Object>>#asCons
    ^ Cons fromObject: self

Cons class>>#fromObject: anObj
    ^ anObj
\end{minted}
and (ii)~the abstract class \Verb|Chain| provides the following class-side messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class>>#bottom
    ^ Bottom new

Chain class>>#item: anObj linker: aBlockClosure
    ^ Knot new
        item: anObj;
        linker: aBlockClosure;
        yourself

Chain class>>#with: anObj
    ^ anObj cons: Chain bottom

Chain class>>#repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
having two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
test case follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \Verb|Chain| object (i)~receives the selector
\Verb|#next:| that carries a \Verb|SmallInteger| object $n$ then it keeps at most
$n$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#next: n
    ^ self

Knot>>#next: n
    ^ n = 1
        ifTrue: [Chain with: item]
        ifFalse: [item cons: [self next next: n - 1]]
\end{minted}

On the other hand, when a \Verb|Chain| object (ii)~receives the selector
\Verb|#contents| it builds a \Verb|LinkedList| object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#contents
    ^ LinkedList new

Knot>>#contents
    ^ self next contents
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot>>#next
    ^ linker value
\end{minted}
Observe that the \Verb|Chain| hierarchy support the protocol provided
by \Verb|ReadStream| objects because of the messages \Verb|#next|, \Verb|#next:|
and \Verb|#contents|.

@@

Recall that a \Verb|Goal| object responds with a \Verb|Chain| that contains either zero,
one, or more substitutions when it receives the message \Verb|#onState:| and it
delegates the task to process those substitutions to \Verb|Runner| objects,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#solutions
    ^ Runner onGoal: self

Runner class>>#onGoal: aGoal
    ^ self new
        goal: aGoal;
        yourself
\end{minted}
which are able to process both (i)~all of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#all
    ^ self run contents
\end{minted}
and (ii)~at most $n$ of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#atMost: anInteger
    ^ (self run next: anInteger) contents
\end{minted}
relying on the unary message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#run
    | states vars selector |
    states := goal onState: Dictionary new asState.
    vars := goal vars ifEmpty: [LinkedList with: Var tautology].
    selector := vars size = 1 ifTrue: #first ifFalse: #yourself.
    ^ states
        collect: [:s | selector
                        value: (s reifier reifyVars: vars)]
\end{minted}
having the responsibilities
\begin{itemize}

>>>>

\documentclass[a4paper,11pt]{article}

@@

% ********************************************************************* 
% Revisions and comments Macros
% ********************************************************************* 
\usepackage{xcolor}
\usepackage[normalem]{ulem} 			% \sout macro
\newboolean{showcomments}
% \setboolean{showcomments}{false}
\setboolean{showcomments}{true}

\ifthenelse{\boolean{showcomments}}
{
	\newcommand{\nb}[3]{
		{\colorbox{#2}{\bfseries\sffamily\scriptsize\textcolor{white}{#1}}}
		{\textcolor{#2}{\sf\small$\blacktriangleright$\textit{#3}$\blacktriangleleft$}}}
	\newcommand{\bnote}[2]{\fbox{\color{blue}\bfseries\sffamily\scriptsize#1}
    	{\color{blue}\sf\small$\blacktriangleright$\textit{#2}$\blacktriangleleft$}}
	\newcommand{\old}[1]{{\color{gray}\sout{#1}}} % old to be removed
	\newcommand{\del}[1]{\old{#1}} % please remove
	\newcommand{\ins}[1]{{\textcolor{blue}{\uline{#1}}}} % please insert	
	\newcommand{\ugh}[1]{{\textcolor{red}{\uwave{#1}}}} % please rephrase	
	\newcommand{\chg}[2]{{\textcolor{red}{\sout{#1}}}{\ra}\textcolor{blue}{\uline{#2}}} % please change
	\newcommand{\here}{\bnote{***}{CONTINUE HERE}} 
	\newcommand{\fix}[1]{\bnote{FIX}{#1}}
}{
	\newcommand{\bnote}[2]{}
	\newcommand{\nb}[3]{}
	\newcommand{\old}[1]{}
	\newcommand{\del}[1]{}
	\newcommand{\ins}[1]{}
	\newcommand{\ugh}[1]{}
	\newcommand{\chg}[2]{}
	\newcommand{\here}{}
	\newcommand{\fix}[1]{}
} 
%---% add your own macros 
\newcommand{\sd}[1]{\nb{Stef}{orange}{#1}}

\newcommand{\ct}[1]{{\textsf{#1}}\xspace}

@@

\sd{As a style point of view I would not put code inside sentence. We should be able to read the text without the code and the code without the text. It makes paragraphs more self contained.}

\sd{Why don't you start with Cons. Then once people understand Cons and infinite stream
you can introduce Goal and state and always example first.}

\sd{Having an diagram showing how an object will be composed will also help}

\subsection{\ct{Goal}s and \ct{State}s}

A \textit{goal} is any object that responds to the message \ct{\#onState:} so
that it receives a substitution and produces an ordered sequence of zero or
more substitutions.

\sd{Show a test. I do not get what is a state? A variable? Why this is call onState:}
\sd{Why this is not \ct{goal withSubstitution: aSubstitution}}

 In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\ct{Dictionary} object fits perfectly \sd{show an example now}; we wrap it \sd{who?} in a \ct{State} object
that also keeps record of an incremental numbering  \ct{birthdate} whose
meaning will be clarified later when we'll describe the  \ct{Fresh}
goal\footnote{put here the reference to its subsection}. An empty substitution
is created by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary new asState
\end{minted}
where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary >> asState
    ^ State new
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> at: aVar put: aValue
    | s |
    s := substitution copy.
    "check that if a key already exists at aVar, then the values must be the same"
    s 
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
maintaining a coherent protocol with respect to the one of \Verb|Dictionary|
objects for extension; moreover, \ct{UnificationError} objects will be
explained in a later section when will talk about the unification process.

\subsection{\ct{Chain}s and \ct{Cons}es}
\sd{As a general comments. You should say to the reader. We will do that. And do piece by piece
Now I discover things just stumbling on them.}

\sd{Example: In this section we will build the infrastructure to represent potentially infinite lists. 
For this will start building a chain. A chain is composed of consed: objects following Lisp tradition. 
We will define first the class Chain as an abstract class with two subclass \ct{Knot} and \ct{Bottom}.
\ct{Bottom} will represent the empty cell of lisp on top of which any list is build. \ct{Knot} represents 
a ...
}

Because the sequence of substitutions may be infinite, we
represent it not as a \ct{LinkedList} object but as a \ct{Chain} object;
as usual, we find \ct{BlockClosure} objects good candidates to represent (a possibly
infinite) computation. Some examples are in order, \sd{ChainTest -> ChainExamples}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest >> collatz: o
    ^ o
        cons: [o even
                ifTrue: [self collatz: o / 2]
                ifFalse: [self collatz: 3 * o + 1]]

ChainTest >> fibs
    ^ self fib: 0 fib: 1

ChainTest >> fib: m fib: n
    ^ m cons: [self fib: n fib: m + n]

ChainTest >> gcd: m gcd: n
    ^ m cons: [n isZero
                ifTrue: [Chain with: m]
                ifFalse: [self gcd: n gcd: m \\ n]]

ChainTest >> nats
    ^ self ints: 0

ChainTest >> ints: i
    ^ i cons: [self ints: i + 1]

ChainTest >> zeroOneTwo
    ^ 0 cons: (1 cons: (2 cons: Chain bottom))
\end{minted}
where any \ct{Object} object respond to the message \ct{cons:} as follows: 
\sd{what is the argument?}
\sd{Separate Cons and Chain explanation}
\sd{why not just started with Cons. Ok we have an object}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> cons: anObj
    ^ anObj consedObject: self
\end{minted}
which dispatches over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> consedObject: car
    ^ Cons car: car asCons cdr: self asCons

BlockClosure >> consedObject: car
    ^ Chain item: car linker: self

Chain >> consedObject: car
    ^ car cons: [self]
\end{minted}
where (i)~\ct{Cons} objects mimic \textit{Lisp's cons cells}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons class >>  car: anObj cdr: anotherObj
    ^ self new
        car: anObj;
        cdr: anotherObj;
        yourself

Object >> asCons
    ^ Cons fromObject: self

Cons class >> fromObject: anObj
    ^ anObj
\end{minted}

\s{I do not get fromObject: implementation. because you do not wrap the object.
So 1 asCons will return 1}
and (ii)~the abstract class \ct{Chain} provides the following class-side messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class >> bottom
    ^ Bottom new

Chain class >> item: anObj linker: aBlockClosure
    ^ Knot new
        item: anObj;
        linker: aBlockClosure;
        yourself

Chain class >> with: anObj
    ^ anObj cons: Chain bottom

Chain class >> repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
having two subclasses \ct{Knot} and \ct{Bottom}. A comprehensive
test case follows \sd{split it and explain.}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest >> testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \ct{Chain} object (i)~receives the message
\ct{next:} that carries a \ct{SmallInteger} object $n$ then it keeps at most
$n$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom >> next: n
    ^ self

Knot >> next: n
    ^ n = 1
        ifTrue: [Chain with: item]
        ifFalse: [item cons: [self next next: n - 1]]
\end{minted}

On the other hand, when a \ct{Chain} object (ii)~receives the message
\ct{#contents} it builds a \ct{LinkedList} object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom >> contents
    ^ LinkedList new

Knot >> contents
    ^ self next contents
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot>>#next
    ^ linker value
\end{minted}
Observe that the \ct{|Chain} hierarchy supports the protocol provided
by \ct{ReadStream} objects because of the messages \ct{#next}, \ct{#next:}
and \ct{#contents}.

@@

Recall \sd{it was not said that way so I do not recall} that a \ct{Goal} object responds with a \ct{Chain} that contains either zero,
one, or more substitutions when it receives the message \ct{#onState:} and it
delegates the task to process those substitutions to \ct{Runner} objects,

\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal >> solutions
    ^ Runner onGoal: self

Runner class >> onGoal: aGoal
    ^ self new
        goal: aGoal;
        yourself
\end{minted}
which are able to process both (i)~all of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> all
    ^ self run contents
\end{minted}
and (ii)~at most $n$ of them with the message \ct{atMost:} below: 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> atMost: anInteger
    ^ (self run next: anInteger) contents
\end{minted}
relying on the unary message \ct{run}

\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#run
    | states vars selector |
    states := goal onState: Dictionary new asState.
    vars := goal vars ifEmpty: [LinkedList with: Var tautology].
    selector := vars size = 1 ifTrue: #first ifFalse: #yourself.
    ^ states
        collect: [:s | selector
                        value: (s reifier reifyVars: vars)]
\end{minted}
having the responsibilities
\begin{itemize}
